---
title: "PERMANOVA"
author: "[Danilo Di Leo](https://github.com/danilodileo)"
date: today
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 2
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    fig-width: 10
    fig-height: 6
    highlight-style: github
    self-contained: true  
editor: visual
---

```{r setup}
#| echo: false
#| cache: false
knitr::opts_chunk$set(echo = TRUE, fig.path = 'figures/', cache = TRUE, fig.width = 10)
ggplot2::theme_set(ggplot2::theme_bw())
```

## Libraries

```{r libraries}
#| message: false
#| warning: false
#| echo: false
#| cache: false
library(tidyverse)
library(vegan)
library(ggrepel)
library(patchwork)
library(gt)
```

## Load data

```{r read-data}
#| cache: true
mt <- read_tsv('./data/mt.tpm.tsv.gz', show_col_types = FALSE) |>
  rename(orf = 1) |>
  pivot_longer(cols = -orf, names_to = "sample", values_to = "tpm_t") |>
  mutate(sample = as.Date(sample))

mg <- read_tsv('./data/mg.tpm.tsv.gz', show_col_types = FALSE) |>
  rename(orf = 1) |>
  pivot_longer(cols = -orf, names_to = "sample", values_to = "tpm_g") |>
  mutate(sample = as.Date(sample))

mp <- read_tsv('./data/harmonized.wide.tsv', show_col_types = FALSE) |>
  rename(orf = 1) |>
  pivot_longer(cols = -orf, names_to = "sample", values_to = "tpm_p") |>
  mutate(sample = as.Date(sample))

annotation <- read_tsv('./data/orf_annotations.tsv.gz', show_col_types = FALSE)
taxonomy   <- read_tsv('./data/orf_taxonomy.tsv.gz', show_col_types = FALSE)
ko <- read_tsv('./data/harmonized.kofamscan.besthit.tsv.gz',
               show_col_types = FALSE) |>
  filter(sign == '*') |>
  rename(orf = 'accno')
```

## Define sample sets

```{r sample-sets}
# ORF pool from metaP
# Note: mp uses harmonized data (TMT + iTRAQ merged via HarmonizerR)
# tpm_p > 0 identifies ORFs present in the harmonized metaP dataset
orfs_mp <- mp |>
  filter(tpm_p > 0) |>
  distinct(orf) |>
  pull(orf)

# Sample sets per pair
samples_mp    <- mp |> distinct(sample) |> pull(sample)
samples_mt    <- mt |> distinct(sample) |> pull(sample)
samples_mg    <- mg |> distinct(sample) |> pull(sample)

samples_mt_mp <- intersect(samples_mt, samples_mp)
samples_mg_mp <- intersect(samples_mg, samples_mp)
samples_all3  <- intersect(samples_mt_mp, samples_mg_mp)

cat("ORFs in metaP pool:     ", length(orfs_mp), "\n")
cat("Campioni metaT ∩ metaP:", length(samples_mt_mp), "\n")
cat("Campioni metaG ∩ metaP:", length(samples_mg_mp), "\n")
cat("Campioni tutti e tre:  ", length(samples_all3),  "\n")
```

## Filter datasets

```{r filter-datasets}
#| cache: true
# metaT — campioni metaT ∩ metaP, solo ORF metaP pool
mt_perm <- mt |>
  filter(sample %in% samples_mt_mp,
         orf %in% orfs_mp)

gc()

# metaG — campioni metaG ∩ metaP, solo ORF metaP pool, solo tpm_g > 0
mg_perm <- mg |>
  filter(sample %in% samples_mg_mp,
         orf %in% orfs_mp,
         tpm_g > 0)

cat("mt_perm:", nrow(mt_perm), "righe |",
    n_distinct(mt_perm$orf), "ORF |",
    n_distinct(mt_perm$sample), "campioni\n")
cat("mg_perm:", nrow(mg_perm), "righe |",
    n_distinct(mg_perm$orf), "ORF |",
    n_distinct(mg_perm$sample), "campioni\n")
```

## Build PA matrices

```{r build-pa}
# Presence/absence matrices
# Rows = samples, columns = ORFs, values = 1 if detected, 0 if not
# Note: metaP uses harmonized data — PA not meaningful for metaP
# Therefore pa_stacked uses only metaT and metaG

# metaT PA — 12 campioni × 4729 ORF
pa_t <- mt_perm |>
  mutate(present = as.integer(tpm_t > 0)) |>
  select(sample, orf, present) |>
  pivot_wider(names_from = orf, values_from = present, values_fill = 0L) |>
  arrange(sample) |>
  column_to_rownames("sample")

# metaG PA — 11 campioni × ORF variabili
pa_g <- mg_perm |>
  mutate(present = as.integer(tpm_g > 0)) |>
  select(sample, orf, present) |>
  pivot_wider(names_from = orf, values_from = present, values_fill = 0L) |>
  arrange(sample) |>
  column_to_rownames("sample")

# Check variazione
cat("=== metaT ===\n")
cat("ORF sempre presenti:", sum(colSums(pa_t) == 12), "\n")
cat("ORF variabili:      ", sum(colSums(pa_t) < 12 & colSums(pa_t) > 0), "\n")
cat("\n=== metaG ===\n")
cat("ORF sempre presenti:", sum(colSums(pa_g) == 11), "\n")
cat("ORF variabili:      ", sum(colSums(pa_g) < 11 & colSums(pa_g) > 0), "\n")
```

## Season helper

```{r season-helper}
get_season <- function(date) {
  month <- as.integer(format(date, "%m"))
  case_when(
    month %in% c(12, 1, 2)  ~ "winter",
    month %in% c(3, 4, 5)   ~ "spring",
    month %in% c(6, 7, 8)   ~ "summer",
    month %in% c(9, 10, 11) ~ "autumn"
  )
}
```

## PCoA — metaT and metaG on metaP ORF pool

```{r pcoa}
#| fig-width: 14
#| fig-height: 6
#| warning: false

make_pcoa_single <- function(pa_mat, layer_name, layer_color) {
  jac  <- vegdist(pa_mat, method = "jaccard", binary = TRUE)
  pcoa <- cmdscale(jac, k = 2, eig = TRUE)
  pct  <- round(pcoa$eig / sum(pcoa$eig[pcoa$eig > 0]) * 100, 1)

  as_tibble(pcoa$points, .name_repair = "minimal") |>
    set_names("PC1", "PC2") |>
    mutate(
      sample = as.Date(rownames(pa_mat)),
      season = factor(get_season(sample),
                      levels = c("winter", "spring", "summer", "autumn"))
    ) |>
    ggplot(aes(x = PC1, y = PC2, shape = season)) +
    geom_point(size = 3.5, alpha = 0.85, color = layer_color) +
    ggrepel::geom_text_repel(
      aes(label = format(sample, "%y-%m")),
      size = 2.5, max.overlaps = 20, color = layer_color
    ) +
    scale_shape_manual(values = c(
      winter = 15, spring = 17, summer = 16, autumn = 18
    )) +
    labs(
      x     = paste0("PC1 (", pct[1], "%)"),
      y     = paste0("PC2 (", pct[2], "%)"),
      shape = "Season",
      title = paste0(layer_name, " — PCoA on metaP ORF pool"),
      subtitle = "Jaccard distance on presence/absence of metaP-matched ORFs"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right")
}

p_mt <- make_pcoa_single(pa_t, "metaT", "#1D9E75")
p_mg <- make_pcoa_single(pa_g, "metaG", "#378ADD")

p_mt + p_mg

# ── PCoA — metaT e metaG insieme, solo ORF metaP pool ────────────────────────

# Allinea colonne tra pa_t e pa_g
all_cols <- sort(unique(c(colnames(pa_t), colnames(pa_g))))

fill_mat <- function(mat, cols) {
  missing <- setdiff(cols, colnames(mat))
  if (length(missing) > 0) mat[missing] <- 0L
  mat[, cols]
}

pa_t_aligned <- fill_mat(pa_t, all_cols)
pa_g_aligned <- fill_mat(pa_g, all_cols)

# Stack metaT + metaG
pa_tg <- rbind(pa_t_aligned, pa_g_aligned)

# Jaccard distance
jac_tg <- vegdist(pa_tg, method = "jaccard", binary = TRUE)

# PCoA
pcoa_tg  <- cmdscale(jac_tg, k = 2, eig = TRUE)
pct_tg   <- round(pcoa_tg$eig / sum(pcoa_tg$eig[pcoa_tg$eig > 0]) * 100, 1)

# Dataframe
pcoa_tg_df <- as_tibble(pcoa_tg$points, .name_repair = "minimal") |>
  set_names("PC1", "PC2") |>
  mutate(
    sample = as.Date(c(rownames(pa_t), rownames(pa_g))),
    layer  = factor(
      c(rep("metaT", nrow(pa_t)), rep("metaG", nrow(pa_g))),
      levels = c("metaT", "metaG")
    ),
    season = factor(get_season(sample),
                    levels = c("winter", "spring", "summer", "autumn"))
  )

# Plot
ggplot(pcoa_tg_df, aes(x = PC1, y = PC2, color = layer, shape = season)) +
  geom_point(size = 3.5, alpha = 0.85) +
   geom_line(
    data = pcoa_tg_df |> filter(sample %in% samples_all3),
    aes(group = sample), color = "grey70", linewidth = 0.4
  ) +
  ggrepel::geom_text_repel(
    aes(label = format(sample, "%y-%m")),
    size = 2.5, max.overlaps = 20
  ) +
  scale_color_manual(values = c(
    "metaT" = "#1D9E75",
    "metaG" = "#378ADD"
  )) +
  scale_shape_manual(values = c(
    winter = 15, spring = 17, summer = 16, autumn = 18
  )) +
  labs(
    x        = paste0("PC1 (", pct_tg[1], "%)"),
    y        = paste0("PC2 (", pct_tg[2], "%)"),
    color    = "Layer",
    shape    = "Season",
    title    = "PCoA — metaT and metaG on metaP ORF pool",
    subtitle = "Jaccard distance on presence/absence of metaP-matched ORFs"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

## PERMANOVA

```{r permanova}
# Note: PERMANOVA run separately for metaT and metaG
# MetaP excluded — harmonized data (HarmonizerR) does not produce 
# meaningful presence/absence values

# Metadata metaT
meta_mt <- tibble(
  sample = as.Date(rownames(pa_t)),
  season = factor(get_season(as.Date(rownames(pa_t))),
                  levels = c("winter", "spring", "summer", "autumn"))
) |>
  left_join(
    mt_perm |>
      group_by(sample) |>
      summarise(n_orfs = sum(tpm_t > 0), .groups = "drop"),
    by = "sample"
  ) |>
  left_join(
    mt_perm |>
      filter(tpm_t > 0) |>
      left_join(taxonomy |> distinct(orf, family), by = "orf") |>
      group_by(sample, family) |>
      summarise(n = n(), .groups = "drop") |>
      group_by(sample) |>
      mutate(prop = n / sum(n)) |>
      filter(family %in% c("Rhodobacteraceae", "Pelagibacteraceae",
                            "Burkholderiaceae", "Cyanobiaceae")) |>
      pivot_wider(names_from = family, values_from = prop,
                  values_fill = 0, id_cols = sample),
    by = "sample"
  ) |>
  replace_na(list(Rhodobacteraceae = 0, Pelagibacteraceae = 0,
                  Burkholderiaceae = 0, Cyanobiaceae = 0)) |>
  column_to_rownames("sample")

# Metadata metaG
meta_mg <- tibble(
  sample = as.Date(rownames(pa_g)),
  season = factor(get_season(as.Date(rownames(pa_g))),
                  levels = c("winter", "spring", "summer", "autumn"))
) |>
  left_join(
    mg_perm |>
      group_by(sample) |>
      summarise(n_orfs = sum(tpm_g > 0), .groups = "drop"),
    by = "sample"
  ) |>
  left_join(
    mg_perm |>
      filter(tpm_g > 0) |>
      left_join(taxonomy |> distinct(orf, family), by = "orf") |>
      group_by(sample, family) |>
      summarise(n = n(), .groups = "drop") |>
      group_by(sample) |>
      mutate(prop = n / sum(n)) |>
      filter(family %in% c("Rhodobacteraceae", "Pelagibacteraceae",
                            "Burkholderiaceae", "Cyanobiaceae")) |>
      pivot_wider(names_from = family, values_from = prop,
                  values_fill = 0, id_cols = sample),
    by = "sample"
  ) |>
  replace_na(list(Rhodobacteraceae = 0, Pelagibacteraceae = 0,
                  Burkholderiaceae = 0, Cyanobiaceae = 0)) |>
  column_to_rownames("sample")

set.seed(42)

cat("=== PERMANOVA metaT ===\n")
adonis2(pa_t ~ season + n_orfs + Rhodobacteraceae + Pelagibacteraceae +
               Burkholderiaceae + Cyanobiaceae,
        data = meta_mt, method = "jaccard",
        binary = TRUE, permutations = 999, by = "margin") |> print()

cat("\n=== PERMANOVA metaG ===\n")
adonis2(pa_g ~ season + n_orfs + Rhodobacteraceae + Pelagibacteraceae +
               Burkholderiaceae + Cyanobiaceae,
        data = meta_mg, method = "jaccard",
        binary = TRUE, permutations = 999, by = "margin") |> print()
```

```{r permanova-table}
# Tabella PERMANOVA results
tibble(
  Layer = c("metaT", "metaT", "metaT", "metaT", "metaT", "metaT",
            "metaG", "metaG", "metaG", "metaG", "metaG", "metaG"),
  Variable = rep(c("Season", "n_orfs", "Rhodobacteraceae", 
                   "Pelagibacteraceae", "Burkholderiaceae", "Cyanobiaceae"), 2),
  R2 = c(0.183, 0.053, 0.061, 0.040, 0.065, 0.046,
         0.142, 0.022, 0.038, 0.022, 0.068, 0.051),
  F_stat = c(1.802, 1.566, 1.805, 1.170, 1.916, 1.348,
             1.506, 0.711, 1.213, 0.702, 2.156, 1.620),
  p_value = c(0.047, 0.128, 0.082, 0.314, 0.079, 0.210,
              0.226, 0.617, 0.335, 0.644, 0.089, 0.226)
) |>
  mutate(
    sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  ) |>
  gt(groupname_col = "Layer") |>
  tab_header(
    title    = "PERMANOVA results — Jaccard distance on metaP ORF pool",
    subtitle = "Marginal effects, 999 permutations"
  ) |>
  cols_label(
    Variable  = "Variable",
    R2        = "R²",
    F_stat    = "F",
    p_value   = "p-value",
    sig       = ""
  ) |>
  fmt_number(columns = c(R2, F_stat), decimals = 3) |>
  fmt_number(columns = p_value, decimals = 3) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = p_value < 0.05
    )
  ) |>
  tab_footnote(
    footnote = "Significance: '***' < 0.001, '**' < 0.01, '*' < 0.05, '.' < 0.1",
    locations = cells_column_labels(columns = sig)
  )
```

The ORF pool detected by metaP represents a genomically stable subset of the bacterioplankton community as genes are present across seasons in the metagenome. For metaT, the composition is varying significantly with season (R² = 0.18, p = 0.047). This indicates that metaproteomics captures proteins encoded by constitutive genomic features while their transcriptional activity fluctuates with environmental conditions. **MetaP captures proteins from genomically stable genes whose transcription can varies seasonally**, examples of these type of proteins are ribosomes!

### Figure 2B - Percentile distribution ORF detected

```{r fig-2b}
#| fig-height: 5
# ── Fig 2B — metaT e metaG su tutti i campioni metaP ─────────────────────────

# metaT — media TPM per ORF, campioni metaT ∩ metaP
mean_tpm_mt <- mt |>
  filter(sample %in% samples_mt_mp) |>
  group_by(orf) |>
  summarise(mean_tpm = mean(tpm_t, na.rm = TRUE), .groups = "drop") |>
  mutate(
    detected_mp = orf %in% orfs_mp,
    layer       = "metaT"
  )

# metaG — media TPM per ORF, campioni metaG ∩ metaP
mean_tpm_mg <- mg |>
  filter(sample %in% samples_mg_mp) |>
  group_by(orf) |>
  summarise(mean_tpm = mean(tpm_g, na.rm = TRUE), .groups = "drop") |>
  mutate(
    detected_mp = orf %in% orfs_mp,
    layer       = "metaG"
  )

# Calcola percentili e proporzione per entrambi
percentile_plot <- bind_rows(mean_tpm_mt, mean_tpm_mg) |>
  group_by(layer) |>
  mutate(
    percentile = cut(
      rank(-mean_tpm) / n(),
      breaks         = seq(0, 1, 0.1),
      labels         = paste0(seq(10, 100, 10), "%"),
      include.lowest = TRUE
    )
  ) |>
  group_by(layer, percentile) |>
  summarise(
    n_total    = n(),
    n_detected = sum(detected_mp),
    prop       = n_detected / n_total,
    .groups    = "drop"
  ) |>
  mutate(layer = factor(layer, levels = c("metaT", "metaG")))

# Plot
ggplot(percentile_plot, aes(x = percentile, y = prop, fill = layer)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(
    "metaT" = "#1D9E75",
    "metaG" = "#378ADD"
  )) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x        = "Abundance percentile (10% bins)",
    y        = "Proportion of ORFs detected in metaP",
    fill     = NULL,
    title    = "MetaP detection bias across abundance percentiles",
    subtitle = "Higher abundance ORFs in metaT and metaG are preferentially detected in metaP"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

### Fig2C - Ranked ORFs metaT and metaG 

```{r fig-2c}
# ── Fig 2C — Ranked ORFs metaT e metaG con metaP highlighted ─────────────────

# Rank ORFs per abbondanza media in metaT
ranked_mt <- mt |>
  filter(sample %in% samples_mt_mp) |>
  group_by(orf) |>
  summarise(mean_tpm = mean(tpm_t, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(mean_tpm)) |>
  mutate(
    rank        = row_number(),
    detected_mp = orf %in% orfs_mp,
    layer       = "metaT"
  )

# Rank ORFs per abbondanza media in metaG
ranked_mg <- mg |>
  filter(sample %in% samples_mg_mp) |>
  group_by(orf) |>
  summarise(mean_tpm = mean(tpm_g, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(mean_tpm)) |>
  mutate(
    rank        = row_number(),
    detected_mp = orf %in% orfs_mp,
    layer       = "metaG"
  )

# Plot
ggplot() +
  geom_line(
    data = ranked_mt |> filter(rank <= 5000),
    aes(x = rank, y = mean_tpm, color = layer),
    linewidth = 0.8
  ) +
  geom_line(
    data = ranked_mg |> filter(rank <= 5000),
    aes(x = rank, y = mean_tpm, color = layer),
    linewidth = 0.8, linetype = "dashed"
  ) +
  geom_point(
    data = ranked_mt |> filter(detected_mp, rank <= 5000),
    aes(x = rank, y = mean_tpm),
    color = "#EF9F27", size = 2, alpha = 0.7
  ) +
  geom_point(
    data = ranked_mg |> filter(detected_mp, rank <= 5000),
    aes(x = rank, y = mean_tpm),
    color = "#EF9F27", size = 2, alpha = 0.7, shape = 17
  ) +
  scale_color_manual(values = c(
    "metaT" = "#1D9E75",
    "metaG" = "#378ADD"
  )) +
  scale_y_log10(labels = scales::comma) +
  labs(
    x        = "ORF rank (by mean abundance)",
    y        = "Mean TPM (log scale)",
    color    = "Layer",
    title    = "ORFs ranked by abundance — metaT and metaG",
    subtitle = "Orange = ORFs detected in metaP · Top 5000 shown"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r}
mp %>%
  left_join(taxonomy, join_by(orf)) %>%
  distinct(genome)
```
